Linearity and tropical mathematics in the Lambda-calculus

### 2023-07-23 CSL24/section1.tex
Commit

In recent years, more and more interest in the programming language community has been directed towards the study of \emph{quantitative} properties of programs like computing the number of computation steps or convergence probabilities, 
as opposed to purely \emph{qualitative} properties like termination or program equivalence. 
Notably, a significant effort has been made to extend, or adapt, well-established qualitative methods, like type systems, relational logics or denotational semantics, to account for quantitative properties. We can mention, for example, 
intersection type systems aimed at capturing time or space resources \cite{decarvalho2018, Accattoli2022} or convergence probabilities \cite{Breuvart2018, PistoneLICS2022},  relational logics to account for probabilistic properties like e.g.~differential privacy \cite{Barthe_2012} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for 
probabilistic \cite{Ehrhard2011, Staton2017} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
The main reason to look for methods relying on (quantitative extensions of) type-theory or denotational semantics is that these approaches yield \emph{modular} and \emph{compositional} techniques, that is, allow one to deduce properties of complex programs from the properties of their constituent parts.

\subsection{Two kinds of quantitative approaches}

Among such quantitative approaches, two different directions have received considerable attention.

On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask   whether they compute functions which do not differ \emph{too much}. This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.

On the other hand, there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the so-called \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}. This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable for the probabilistic setting \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.

@@

Tropical mathematics was introduced in the seventies by the Brazilian mathematician Imre Simon \cite{Simon} as an alternative approach to algebra and geometry where the usual ring structure of numbers based on addition and multiplication is replaced by the semiring structure given, respectively, by ``$\min$'' and ``$+$''.
%
%
% interpreting the usual ``$\times$'' and  ``$+$'' operations by  ``$+$'' by ``$\min$''. It can thus be seen as a sort of ``logarithmic'' version of usual geometry (this idea can be made precise via the so-called \emph{Maslov deformation} \cite{}).
%Tropical mathematics is a form of \emph{idempotent} mathematics, since the role of addition is 
%played by the idempotent operation $\min$.
For instance, the polynomial $p(x,y)=x^{2}+xy^{2}+y^{3}$, when interpreted over the tropical semiring, translates as the piecewise linear function
$
\varphi(x,y)=\min\{2x, x+2y, 3y\}
$.

%This is not a \emph{ad-hoc} setting: 
In the last decades, tropical geometry evolved into a vast and rich research domain, providing a combinatorial counterpart of usual algebraic geometry, with important connections with optimisation theory \cite{Sturmfelds}.
Computationally speaking, working with $\min$ and $+$ is generally easier than working with standard addition and multiplication; for instance, the fundamental (and generally intractable) problem of finding the roots of a polynomial admits a \emph{linear time} algorithm in the tropical case (and, moreover,  the tropical roots can be used to approximate the actual roots \cite{Noferini2015}).
The computational nature of tropical notions explains why these are so widely applied in computer science, notably for convex analysis and machine learning (see \cite{Maragos2021} for a recent survey).

Coming back to our discussion on program semantics, tropical geometry might seem to provide precisely what look for, as it turns the monomials $x^{n}$ into the Lipschitz map  $n\cdot x$.
At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail. Furthermore, connections between tropical linear algebra and metric spaces have also been observed \cite{Fuji} within the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}.
However, a thorough investigation of the interpretation of the $\lambda$-calculus within tropical mathematics has not yet been undertaken.

@@

\subsection{Contributions}

>>>>

In recent years, more and more interest in the programming language community has been directed towards the study of \emph{quantitative} properties of programs like computing the number of computation steps or convergence probabilities, 
as opposed to purely \emph{qualitative} properties like termination or program equivalence. 
Notably, a significant effort has been made to extend, or adapt, well-established qualitative methods, like type systems, relational logics or denotational semantics, to account for quantitative properties. We can mention, for example, 
intersection type systems aimed at capturing time or space resources \cite{decarvalho2018, Accattoli2022} or convergence probabilities \cite{Breuvart2018, PistoneLICS2022},  relational logics to account for probabilistic properties like e.g.~differential privacy \cite{Barthe_2012} or metric preservation \cite{Reed2010, dallago}, as well as the study of denotational models for 
probabilistic \cite{Ehrhard2011, Staton2017} or differential \cite{difflambda} extensions of the $\lambda$-calculus. 
The main reason to look for methods relying on (quantitative extensions of) type-theory or denotational semantics is that these approaches yield \emph{modular} and \emph{compositional} techniques, that is, allow one to deduce properties of complex programs from the properties of their constituent parts.   
Among such quantitative approaches, two different directions have received considerable attention, which we recall below.

%\subsection{Two kinds of quantitative approaches}\label{section2}

\input{Section2}

%On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask   whether they compute functions which do not differ \emph{too much}. This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.

%On the other hand, there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the so-called \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}. This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable for the probabilistic setting \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.

@@

Tropical mathematics was introduced in the seventies by the Brazilian mathematician Imre Simon \cite{Simon} as an alternative approach to algebra and geometry where the usual ring structure of numbers based on addition and multiplication is replaced by the semiring structure given, respectively, by ``$\min$'' and ``$+$''.
%
%
% interpreting the usual ``$\times$'' and  ``$+$'' operations by  ``$+$'' by ``$\min$''. It can thus be seen as a sort of ``logarithmic'' version of usual geometry (this idea can be made precise via the so-called \emph{Maslov deformation} \cite{}).
%Tropical mathematics is a form of \emph{idempotent} mathematics, since the role of addition is 
%played by the idempotent operation $\min$.
For instance, the polynomial $p(x,y)=x^{2}+xy^{2}+y^{3}$, when interpreted over the tropical semiring, translates as the piecewise linear function
$
\varphi(x,y)=\min\{2x, x+2y, 3y\}
$.
%This is not a \emph{ad-hoc} setting: 
In the last decades, tropical geometry evolved into a vast and rich research domain, providing a combinatorial counterpart of usual algebraic geometry, with important connections with optimisation theory \cite{Sturmfelds}.
Computationally speaking, working with $\min$ and $+$ is generally easier than working with standard addition and multiplication; for instance, the fundamental (and generally intractable) problem of finding the roots of a polynomial admits a \emph{linear time} algorithm in the tropical case (and, moreover,  the tropical roots can be used to approximate the actual roots \cite{Noferini2015}).
The computational nature of tropical notions explains why these are so widely applied in computer science, notably for convex analysis and machine learning (see \cite{Maragos2021} for a recent survey).

\subsection{This paper and our contributions}

Coming back to our discussion on program semantics %tropical geometry might seem to provide precisely what look for, as it turns the monomials $x^{n}$ into the Lipschitz map  $n\cdot x$.
the metric approach usually yields Lipschitz conditions for affine programs, but what about ``unbounded'', simply typed $\lambda$-calculus (the ordinary one)?
In such case, one might lose the Lipschitz property, so the programs might be \emph{too} sensitive to errors. 
%For instance, while the functions $M_{k}=\lambda x. k\cdot x: \BB R\to \BB R$ are all Lipschitz-continuous, with Lipschitz constant $k$, the program $M=\lambda x.x^{2}$ obtained by ``duplicating'' $x$ is not Lipschitz anymore: $M$ is, so to say, \emph{too} sensitive to errors. 
It is indeed well-known that the category $\Met$ is \emph{not} Cartesian closed, so it is not a model of $\STLC$ (yet, several Cartesian closed \emph{sub-}categories of $\Met$ do exist, see e.g.~\cite{Clementino2006, PistoneFSCD2022}).
%Still, one might observe that the program $M$ above is actually \emph{locally} Lipschitz-continuous (i.e.~Lipschitz over any open ball).
However, one can still consider Cartesian closed categories of \emph{locally} Lipschitz maps  (i.e.~Lipschitz over any open ball), see e.g.\ \cite{Ehrhard2011, PistoneLICS}, and a new example will be considered in this paper.

On the other hand, since the resource approach decomposes, through the Taylor formula, an unbounded application as a limit of bounded ones, one might well ask whether it could be possible to see this formula as a decomposing the associated locally Lipschitz map by a limit of the associated Lipschitz ones, thus bridging the metric and differential approaches.  
Here, a natural direction to look for is the \emph{relational semantics}, since strictly related with the Taylor expansion $\STDLC$ {\color{red}CITE DeCarvalho ecc}. 
However, in this semantics, terms with bounded applications correspond to \emph{polynomials}, i.e.~to non-Lipschitz functions.

Yet, what if these polynomials were tropical ones, i.e.~piecewise linear functions? This way, the Taylor expansion could really be interpreted as a decomposition of $\lambda$-terms via limits (indeed, $\inf$s) of Lipschitz maps: unbounded term application could be seen as a limit of \emph{more and more sensitive} operations.

%This viewpoint, that we develop in the following sections, not only suggests the application of optimization methods based on tropical mathematics to the study of the $\lambda$-calculus and its quantitative extensions, but it scales to a more abstract level, leading to introduce a differential operator for continuous functors between \emph{generalized} metric spaces (in the sense of \cite{Lawvere1973}).
%
%At this point, 
%by interpreting such polynomials over the tropical semiring 
%
%Our question can thus be reformulated as follows: can we make the relational semantics \emph{Lipschitz}, hence amenable to metric and sensitivity analysis? The goal of this paper is to show that, by appealing to tropical mathematics, this is indeed possible and leads to the somehow unespected discovery of a bridge between the metric and differential study to higher-order programs.

At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail. Furthermore, connections between tropical linear algebra and metric spaces have also been observed \cite{Fuji} within the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}.
However, a thorough investigation of the interpretation of the $\lambda$-calculus within tropical mathematics has not yet been undertaken.

%\subsection{Contributions}

@@

### 2023-07-23 CSL24/section2.tex
Commit

\subsection{Metric approach: bounded $\lambda$-calculus in a nutshell}

In many situations (e.g.~when dealing with computationally difficult problems) one does not look for algorithms to compute a function \emph{exactly}, but rather to approximate it (in an efficient way) within some error bound. In other common situations (e.g.~in differential privacy \cite{Alvim2011, Reed2010}) one needs to verify that an algorithm is not \emph{too sensitive} to errors, that is, a small error in the input will produce a comparably small error in the output.

%In all these cases, it is common to consider forms of denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which account for differences in behavior. 
A fundamental insight coming from this line of work is that \emph{affine} programs, i.e.~programs that may use their input at most once, correspond to \emph{non-expansive} (or $1$-Lipschitz) maps, that is, to functions $f$ for which the distances
$d(f(x),f(y))$ produced in output are bounded by the distances $d(x,y)$ in input. 
A more formal way of stating this observation is that the category $\Met$ of pseudo-metric spaces and non-expansive maps provides a model of the \emph{linear} simply typed $\lambda$-calculus, being a \emph{symmetric monoidal closed} category, and in fact it also models {affine} terms (since the cartesian and monoidal units coincide).

As observed in \cite{Reed2010, Gaboardi2017}, the metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplications. The fundamental intuition is that a program duplicating its input $K$ times will give rise to a $K$-\emph{Lipschitz} map.
For instance, the higher-order program $M=\lambda f.\lambda x.f(f(x))$, which duplicates the functional input $f$, yields a $2$-Lipschitz map between the metric space $\BB R\multimap \BB R$ of non-expansive real functions and itself: if $f,g$ are two non-expansive maps differing by at most $\epsilon$ (i.e.~for which $|f(x)-g(x)|\leq \epsilon$ holds for all $x\in \BB R$), then the application of $M$ to $f$ and $g$ will produce two maps differing by at most $2\epsilon$. 
By observing that a $r$-Lipschitz map between metric spaces $X$ and $Y$ is the same as a non-expansive map between the \emph{re-scaled} space $r\cdot X$ (i.e.~with distance $d_{r}(x,y)=r\cdot d(x,y)$) and $Y$, the program $M$ above 
can thus be interpreted as a non-expansive map from $2\cdot(\BB R\multimap \BB R)$ to $\BB R\multimap \BB R$.

These observations have led to the study of linear $\lambda$-calculi with \emph{graded} exponentials, like $\mathsf{Fuzz}$ \cite{Reed2010}, inspired from Girard's Bounded Linear Logic \cite{Girard92tcs}, which have been applied to the study of differential privacy \cite{Gaboardi2013, Gaboardi2017}. The types of such systems are defined by combining linear constructors with a \emph{graded linear exponential comonad} $!_{r}(-)$ \cite{Katsumata2018}.
The language $\BSTLC$ we consider is a simplified version of the language $\mathrm{Fuzz}$ \cite{Reed2010}. 
The terms are as for the $\STLC$, the types are $A::= o \ \mid  \ !_{n}A \multimap A$ and type judgements are of the form $\Gamma \vdash M:A$, where a context $\Gamma$ is a set of declarations of the form $x :_{n}A$, with $n\in \mathbb N$.
The typing rules are given in Fig.~\ref{fig:rules}, where $\Gamma+\Delta$ is defined letting $(\Gamma, x:_{m} A)+( \Delta, x:_{n} A) =  (\Gamma+\Delta), x:_{m+n}A$, and $m\Gamma$ is made all $x:_{mn}A$ for $(x:_{n}A) \in \Gamma$.  
The axiom is $x:_{1}A\vdash x: A$.
The main feature of this language is that if $\vdash \lambda x.M:\,!_nA\multimap B$, then $x$ will be duplicated exactly $n$ times in the reduction to the normal form. %, i.e.\ $\vdash \lambda x.M:\,!_nA\multimap B$ is a program which performs $n$ duplications.
E.g., $\vdash_{\BSTLC} \lambda {\color{red}z}.\left( \lambda x{\color{green}y}. yxx \right)z : \, !_{\color{red}2} o\multimap !_{\color{green}1}(!_{\color{violet}1} o \multimap !_{\color{blue}1} o \multimap o) \multimap o$.

The $\BSTLC$ can be interpreted in a symmetric monoidal closed category (SMCC) equipped with a \emph{$\mathbb N$-graded linear exponential comonad} \cite{Katsumata2018}.
%Observe that one can always type an affine term like e.g.~$\lambda xy.x$ with a linear type $A\multimap B\multimap A$. Instead, a term like $\lambda xy.x(xy)$ containing two occurrences of $x$ cannot be given the linear type $(A\multimap A)\multimap (A\multimap A)$ but a type of the form
%$!_{2}(A\multimap A)\multimap (A\multimap A)$.

@@

\subsection{Resource approach: differential $\lambda$-calculus in a nutshell}

A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} \cite{difflambda} (and differential linear logic \cite{dill}) and its categorical models. 
The key ingredient is a \emph{differential operator} $\Der$,  added to the usual syntax of the $\lambda$-calculus. The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Diff{M}{N}$, still of type $A\to B$, corresponds to the \emph{linear application} of $M$ to $N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Diff{M}{N}$ still has type $A\to B$, since $M$ might need \emph{other} copies of an input of type $A$).

The syntax of the differential $\lambda$-calculus $\STDLC$ (e.g.\ \cite[Section 3]{Manzo2010}), is given by the \emph{terms} $M$ and the \emph{sums} $\mathbb T$, mutually generated by: $M::= x\mid \lambda x.M \mid M\mathbb T \mid \Diff{M}{M}$ and $\mathbb T::= 0 \mid M \mid M+\mathbb T$,
quotiented by: $\alpha$-equivalence; $+,0$ form a commutative monoid on the set of sums; %, i.e.\ commutativity and associativity of $+$ and neutrality of $0$ w.r.t.\ $+$;
linearity of $\lam x.(\_)$, $\Diff{\_}{\_}$ and $(\_)\mathbb T$ (but \emph{not} of $M(\_)$); irrelevance of the order of consecutive $\Diff{\_}{\_}$.
%Remark that $M(\_)$ is \emph{not} set to be linear: $\lambda x.0=0\mathbb T=\Diff{0}{N}=\Diff{M}{0}=0$ but $M0\neq0$ in general.
%This is crucial for the definition of the Taylor expansion.
We follow the tradition of quotienting also for the idempotency of $+$.
%Sums are, then, just \emph{finite} sets of terms.
Simple types are $A::= o\mid A\to A$.
The typing system is in Figure~\ref{fig:rules}, where a context $\Gamma$ is a list of typed variable declarations.
The axioms are $\Gamma, x:A \vdash x: A$ and $\Gamma\vdash 0:A$.

%In $\BSTLC$ the typing system handles duplications; in $\STDLC$ the syntax with its operational semantics (that we do not give) does it.
Writing $\Der^2[\_,(\_)^2]$ as a shortcut for $\Der[\Der[\_,\_],\_]$ and $\Der^1[\_,(\_)^1]$ for $\Diff{\_}{\_}$, the analogue of the previous $\BSTLC$-term is $\vdash_{\STDLC} \lambda {\color{red}z}. \Der^{\color{red}2}[
	\lambda x{\color{green}y}.
		\Der^{\color{violet}1} [
				\Der^{\color{blue}1} [y, x^{\color{blue}1}]
        0, x^{\color{violet}1}
	]0
, z^{\color{red}2}]0
: {\color{red}o}\to ({\color{green}o \to o \to o}) \to o$.
%Here we wrote $\Der^2[\_]\cdot (\_)^2$ as a shortcut for $\Der[\Der[\_]\cdot (\_)]\cdot (\_)$ and $\Der^1[\_]\cdot (\_)^1$ for $\Der[\_]\cdot (\_)$.
In particular, if the \emph{multiplicities} of the arguments (the colored exponents) do not exactly match the number of duplications, e.g.\ in $\vdash_{\STDLC} \lambda z. \Der^{\color{red}3}[
	\lambda xy.
		\Diff{
				\Diff{y}{x}
		0}{x}
	0
, z^{\color{red}3}]0
: o\to (o \to o \to o) \to o$, then the term reduces to the empty sum $0$ (representing an \emph{error}).

>>>>

\paragraph*{Metric approach: bounded $\lambda$-calculus in a nutshell}

On the one hand one there is the approach of \emph{program metrics} \cite{Reed2010, Gaboardi2017, Gabo2019} and \emph{quantitative equational theories} \cite{Plotk}: when considering probabilistic or approximate computation, rather than asking whether two programs compute \emph{the same} function, it makes more sense to ask whether they compute functions which do not differ \emph{too much} by controlling the propagation of an error during the computation.
This has motivated the study of denotational frameworks in which types are endowed with a metric, measuring similarity of behavior; this approach has found  applications in e.g.~differential privacy \cite{Reed2010} and coinductive methods \cite{Bonchi2018}, and was recently extended to account for the full $\lambda$-calculus \cite{Geoffroy2020, PistoneLICS, PistoneFSCD2022}.
%As already mentioned, in many situations (e.g.~when dealing with computationally difficult problems, or in differential privacy \cite{Alvim2011, Reed2010}), one %does not look for algorithms to compute a function \emph{exactly}, but rather to approximate it (in an efficient way) within some error bound. In other common situations (e.g.~in differential privacy \cite{Alvim2011, Reed2010}) one needs to verify that an algorithm is not \emph{too sensitive} to errors, that is, a small error in the input will produce a comparably small error in the output. 
%In all these cases, it is common to consider forms of denotational semantics in which types are endowed with a \emph{behavioral metric}, that is, a metric on programs which account for differences in behavior. 
A fundamental insight coming from this line of work is that \emph{affine} programs, i.e.~programs that may use their input at most once, correspond to \emph{non-expansive} (or $1$-Lipschitz) maps, that is, to functions $f$ for which the distances
$d(f(x),f(y))$ produced in output are bounded by the distances $d(x,y)$ in input. 
This is due to the fact that the category $\Met$ of pseudo-metric spaces and non-expansive maps provides a model of the %\emph{linear} simply typed $\lambda$-calculus, being a \emph{symmetric monoidal closed} category, and in fact it also models 
\emph{affine} $\lambda$-calculus (it is symmetric monoidal closed, cartesian and monoidal units coincide). 
The metric approach is not restricted to affine programs, but can be extended to programs with \emph{bounded} duplications \cite{Reed2010, Gaboardi2017}: the idea is that a program duplicating its input $K$ times will give rise to a $K$-\emph{Lipschitz} map.
For instance, the higher-order program $M=\lambda f.\lambda x.f(f(x))$, which duplicates the function input $f$, yields a $2$-Lipschitz map between the metric space $\BB R\multimap \BB R$ of non-expansive real functions and itself. %: if $f,g$ are two non-expansive maps differing by at most $\epsilon$ (i.e.~for which $|f(x)-g(x)|\leq \epsilon$ holds for all $x\in \BB R$), then the application of $M$ to $f$ and $g$ will produce two maps differing by at most $2\epsilon$. 
By observing that a $r$-Lipschitz map between metric spaces $X$ and $Y$ is the same as a non-expansive map between the \emph{re-scaled} space $rX:=(X,%d_{r}(x,y)=
rd(x,y))$ and $Y$, the program $M$ above 
can thus be interpreted as a non-expansive map from $2(\BB R\multimap \BB R)$ to $\BB R\multimap \BB R$.

These observations have led to the study of linear $\lambda$-calculi with \emph{graded} exponentials, like $\mathsf{Fuzz}$ \cite{Reed2010}, inspired from Girard's Bounded Linear Logic \cite{Girard92tcs}, which have been applied to the study of differential privacy \cite{Gaboardi2013, Gaboardi2017}. The types of such systems are defined by combining linear constructors with a \emph{graded linear exponential comonad} $!_{r}(-)$ \cite{Katsumata2018}.
The language $\BSTLC$ we consider is a simplified version of the language $\mathrm{Fuzz}$ \cite{Reed2010}. 
The terms are as for the $\STLC$, the types are $A::= * \ \mid  \ !_{n}A \multimap A$ and type judgements are of the form $\Gamma \vdash M:A$, where a context $\Gamma$ is a set of declarations of the form $x :_{n}A$, with $n\in \mathbb N$.
The typing rules are given in Fig.~\ref{fig:rules}, where $\Gamma+\Delta$ is defined letting $(\Gamma, x:_{m} A)+( \Delta, x:_{n} A) =  (\Gamma+\Delta), x:_{m+n}A$, and $m\Gamma$ is made all $x:_{mn}A$ for $(x:_{n}A) \in \Gamma$.  
The axiom is $x:_{1}A\vdash x: A$.
The main feature of this language is that if $\vdash \lambda x.M:\,!_nA\multimap B$, then $x$ will be duplicated exactly $n$ times in the reduction to the normal form. %, i.e.\ $\vdash \lambda x.M:\,!_nA\multimap B$ is a program which performs $n$ duplications.
E.g., $\vdash_{\BSTLC} \lambda {\color{red}z}.\left( \lambda x{\color{green}y}. yxx \right)z : \, !_{\color{red}2} *\multimap !_{\color{green}1}(!_{\color{violet}1} * \multimap !_{\color{blue}1} * \multimap *) \multimap *$.
The $\BSTLC$ can be interpreted in a symmetric monoidal closed category (SMCC) equipped with a \emph{$\mathbb N$-graded linear exponential comonad} \cite{Katsumata2018}.
%Observe that one can always type an affine term like e.g.~$\lambda xy.x$ with a linear type $A\multimap B\multimap A$. Instead, a term like $\lambda xy.x(xy)$ containing two occurrences of $x$ cannot be given the linear type $(A\multimap A)\multimap (A\multimap A)$ but a type of the form
%$!_{2}(A\multimap A)\multimap (A\multimap A)$.

@@

\paragraph*{Resource approach: differential $\lambda$-calculus in a nutshell}

On the other hand, there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the so-called \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}. This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable for the probabilistic setting \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.

%A different family of approaches to linearity and duplication arises from the study of the \emph{differential $\lambda$-calculus} \cite{difflambda} (and differential linear logic \cite{dill}) and its categorical models. 
The key ingredient is a \emph{differential operator} $\Der$,  added to the usual syntax of the $\lambda$-calculus. The intuition is that, given $M$ of type $A\to B$ and $N$ of type $A$, the program $\Diff{M}{N}$, still of type $A\to B$, corresponds to the \emph{linear application} of $M$ to $N$: this means that $N$ is passed to $M$ so that the latter may use it exactly once (this is why $\Diff{M}{N}$ still has type $A\to B$, since $M$ might need \emph{other} copies of an input of type $A$).

The syntax of the differential $\lambda$-calculus $\STDLC$ (e.g.\ \cite[Section 3]{Manzo2010}), is given by the \emph{terms} $M$ and the \emph{sums} $\mathbb T$, mutually generated by: $M::= x\mid \lambda x.M \mid M\mathbb T \mid \Diff{M}{M}$ and $\mathbb T::= 0 \mid M \mid M+\mathbb T$,
quotiented by: $\alpha$-equivalence; $+,0$ form a commutative monoid on the set of sums; %, i.e.\ commutativity and associativity of $+$ and neutrality of $0$ w.r.t.\ $+$;
linearity of $\lam x.(\_)$, $\Diff{\_}{\_}$ and $(\_)\mathbb T$ (but \emph{not} of $M(\_)$); irrelevance of the order of consecutive $\Diff{\_}{\_}$.
%Remark that $M(\_)$ is \emph{not} set to be linear: $\lambda x.0=0\mathbb T=\Diff{0}{N}=\Diff{M}{0}=0$ but $M0\neq0$ in general.
%This is crucial for the definition of the Taylor expansion.
We follow the tradition of quotienting also for the idempotency of $+$.
%Sums are, then, just \emph{finite} sets of terms.
Simple types are $A::= *\mid A\to A$.
The typing system is in Figure~\ref{fig:rules}, where a context $\Gamma$ is a list of typed variable declarations.
The axioms are $\Gamma, x:A \vdash x: A$ and $\Gamma\vdash 0:A$.

%In $\BSTLC$ the typing system handles duplications; in $\STDLC$ the syntax with its operational semantics (that we do not give) does it.
Writing $\Der^2[\_,(\_)^2]$ as a shortcut for $\Der[\Der[\_,\_],\_]$ and $\Der^1[\_,(\_)^1]$ for $\Diff{\_}{\_}$, the analogue of the previous $\BSTLC$-term is $\vdash_{\STDLC} \lambda {\color{red}z}. \Der^{\color{red}2}[
	\lambda x{\color{green}y}.
		\Der^{\color{violet}1} [
				\Der^{\color{blue}1} [y, x^{\color{blue}1}]
        0, x^{\color{violet}1}
	]0
, z^{\color{red}2}]0
: {\color{red}*}\to ({\color{green}* \to * \to *}) \to *$.
%Here we wrote $\Der^2[\_]\cdot (\_)^2$ as a shortcut for $\Der[\Der[\_]\cdot (\_)]\cdot (\_)$ and $\Der^1[\_]\cdot (\_)^1$ for $\Der[\_]\cdot (\_)$.
In particular, if the \emph{multiplicities} of the arguments (the colored exponents) do not exactly match the number of duplications, e.g.\ in $\vdash_{\STDLC} \lambda z. \Der^{\color{red}3}[
	\lambda xy.
		\Diff{
				\Diff{y}{x}
		0}{x}
	0
, z^{\color{red}3}]0
: *\to (* \to * \to *) \to *$, then the term reduces to the empty sum $0$ (representing an \emph{error}).

### 2023-07-23 CSL24/section3A.tex
Commit

The linear simply typed $\lam$-calculus is a restriction of the ordinary $\lam$-calculus in which programs, following the linear logic mantra, can only use their arguments \emph{exactly once} (and indeed this calculus can be embedded in its \emph{intuitionistic multiplicative} fragment \emph{IMLL}).
%More precisely, the linear $\lam$-calculus is obtained from the ordinary one by adding the constraint that each $\lam$-abstracted variable appears exactly once in the scope of the $\lam$-abstraction.
%In order to frame it in a category $\C C$, one needs a symmetric tensor product $\otimes$ together with internal hom-set objects $X\multimap Y$ s.t.\ the evaluation and curry maps yield a \emph{symmetric monoidal adjunction}: $\C C(Z\otimes X, Y) \simeq \C C (Z, X\multimap Y)$.
It can be interpreted in any Symmetric Monoidal Closed Category (SMCC).
Now, \cite[Section III.A]{Manzo2013} immediately gives that $\LREL$ is a SMCC (actually, even a model of IMLL), where its monoidal product $\otimes$ acts on the objects as the Cartesian product of sets, and as the \emph{Kronecker product}
 %{(\color{red} Ref?)} 
 of matrices on morphisms.
Its internal hom $X\multimap Y$ is also given by the Cartesian product $X\times Y$ of sets, with the usual evaluation and curry maps.

>>>>

The linear simply typed $\lam$-calculus is a restriction of the ordinary $\lam$-calculus in which programs, following the linear logic mantra, can only use their arguments \emph{exactly once} (and indeed this calculus can be embedded in its \emph{intuitionistic multiplicative} fragment \emph{IMLL}).
%More precisely, the linear $\lam$-calculus is obtained from the ordinary one by adding the constraint that each $\lam$-abstracted variable appears exactly once in the scope of the $\lam$-abstraction.
%In order to frame it in a category $\C C$, one needs a symmetric tensor product $\otimes$ together with internal hom-set objects $X\multimap Y$ s.t.\ the evaluation and curry maps yield a \emph{symmetric monoidal adjunction}: $\C C(Z\otimes X, Y) \simeq \C C (Z, X\multimap Y)$.
It can be interpreted in any Symmetric Monoidal Closed Category (SMCC).
Now, \cite[Section III.A]{Manzo2013} immediately gives that $\LREL$ is a SMCC (actually, even a model of IMLL), where its monoidal product $\otimes$ acts on the objects as the Cartesian product of sets, %and as the \emph{Kronecker product} of matrices on morphisms.
 %{(\color{red} Ref?)} 
 and its internal hom $X\multimap Y$ is also given by the Cartesian product $X\times Y$ of sets, with the usual evaluation and curry maps.
